State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-05-10 analysis/paper/paper.tex
reknit word and html versions, and pdf in order to keep the .tex file

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources},
  pdfauthor={Kevin E. See1,, Ryan N. Kinzer2, and Mike W. Ackerman1},
  pdfkeywords={escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model},
  pdfborder={0 0 0},
  breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-2}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

@@

\title{State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources}
\author{Kevin E. See\textsuperscript{1,*}, Ryan N. Kinzer\textsuperscript{2}, and Mike W. Ackerman\textsuperscript{1}}
\date{August 18, 2020}

\begin{document}
\maketitle
\begin{abstract}
This is the abstract
\end{abstract}

\textsuperscript{1} Biomark, Inc.~705 South 8th St., Boise, Idaho, 83702, USA\\
\textsuperscript{2} Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA

@@

\textbf{Possible journals:}

First Choice:

\begin{itemize}
\tightlist
\item
  \textbf{\href{https://afspubs.onlinelibrary.wiley.com/journal/15488675}{North American Journal of Fisheries Management}}
\end{itemize}

Second Choice:

\begin{itemize}
\tightlist
\item
  \href{https://afspubs.onlinelibrary.wiley.com/journal/15488659}{Transactions of the American Fisheries Society}
\end{itemize}

Others:

\begin{itemize}
\item
  \href{http://www.nrcresearchpress.com/page/cjfas/editors}{Canadian Journal of Fisheries and Aquatic Sciences}
\item
  \href{https://www.journals.elsevier.com/fisheries-research}{Fisheries Research}
\item
  \href{http://fishbull.noaa.gov/}{Fishery Bulletin}
\item
  \href{http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html}{Fish and Fisheries}
\end{itemize}

@@

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.~Bue et al. (1998)). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin {[}wild, hatchery{]}; Hess et al. (2014); Steinhorst et al. (2017)) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics (McElhany et al. 2000), and facilitate assessments of population viability and extinction risk (Ford et al. 2015; Williams et al. 2016) and climate change vulerability (Crozier et al. 2019).

Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats (Nehlsen et al. 1991; McClure et al. 2003; Ford et al. 2015). As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (\textbf{Snake Basin Harvest Plans?}). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (\textbf{could cite US v OR Management Agreement and Snake Basin Harvest Plans?}).

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window (Camacho et al. 2017 p. @Steinhorst2017). Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (\textbf{citation?}).

Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live (i.e., in-person) window counts only occur from April through October each year. Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour. Counts are then expanded to provide an estimate for the hour (USACE 2015). From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts (Hatch et al. 1994). Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted. Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again (Boggs et al. 2004). Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement (Dauble and Mueller 2000). Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (\textbf{citation?})

In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach (Royle and Dorazio 2008). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the \textbf{ST}ate-space \textbf{A}dult \textbf{D}am \textbf{E}scapement \textbf{M}odel (STADEM) package for the statistical software R (R Core Team 2020), and is available for downloaded at \url{https://github.com/KevinSee/STADEM}. To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty. Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.

@@

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours (Hatch et al. 1994). Video monitoring occured during the beginning and tail ends of the adult runs (March 1 -- March 31 and November 1 -- December) for 10 hours per day (0600 -- 1600 hours). Direct visual monitoring occured during peak run times (April 1 -- October 31) for 16 hours per day (0400 -- 2000 hours) (USACE 2015). During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) (USACE 2015). Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, \url{www.cbr.washington.edu/dart/query/adult_daily}, using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam (Ogden 2016a). The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin {[}wild, hatchery{]}, genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., Camacho et al. (2019), Steinhorst et al. (2017)). The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns. Trap sample rates were typically 10-25\%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age (Wright et al. 2015), tissue samples were used to determine sex and for genetic stock identification analysis (e.g., Hargrove et al. (2019)). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game (Camacho et al. 2019).

\hypertarget{pit-tag-data}{%
\subsubsection{PIT tag Data}\label{pit-tag-data}}

The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data was provided through DART and the adult ladder PIT tag query; \url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}.

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The ``mark'' group included all tags detected in the adult trap and the ``capture'' group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.

The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis. The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach (Royle and Dorazio 2008) implemented in the STADEM package for the R statistical software (R Core Team 2020). We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \ref{fig:stadem-examp-fig}. Additional model details can be found in \protect\hyperlink{append1}{Appendix A}. The STADEM package is available from the primary author at a \url{https://github.com/KevinSee/STADEM}, and requires the use of the JAGS software (Plummer 2019) for Bayesian inference.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources},
  pdfauthor={Kevin E. See1,, Ryan N. Kinzer2, and Michael W. Ackerman1},
  pdfkeywords={escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

@@

\title{State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources}
\author{Kevin E. See\textsuperscript{1,*}, Ryan N. Kinzer\textsuperscript{2}, and Michael W. Ackerman\textsuperscript{1}}
\date{May 10, 2021}

\begin{document}
\maketitle
\begin{abstract}
Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon, \emph{Oncorhynchus tshawytscha}, and steelhead, \emph{O. mykiss}, from spawn years 2010-2019.
\end{abstract}

\textsuperscript{1} Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA\\
\textsuperscript{2} Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA

@@

@@

Fish escapement refers to the number of adults that survive juvenile and subadult rearing, escape harvest, and achieve a size and age to potentially spawn (e.g., Bue et al. 1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their natal habitat to spawn (e.g., Woody 2004; Jasper et al. 2018). Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups (e.g., by stock, population, age, origin {[}wild, hatchery{]}, Hess et al. 2014; Steinhorst et al. 2017; Camacho et al. 2018) which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics (McElhany et al. 2000) and facilitate assessments of population viability, extinction risk (Northwest Fisheries Science Center 2015; Williams et al. 2016) and climate change vulnerability (Crozier et al. 2019).

Populations of Chinook Salmon \emph{Oncorhynchus tshawytscha} and anadromous steelhead trout \emph{O. mykiss} in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats (Nehlsen et al. 1991; McClure et al. 2003; Northwest Fisheries Science Center 2015). As a result, most Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam {[}LGR{]}), is monitored at LGR located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR (NPCC 2014; Northwest Fisheries Science Center 2015; National Marine Fisheries Service 2019). As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR (U.S. v Oregon 2018).

The majority of sp/sum Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on LGR before migrating upriver. Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window (Steinhorst et al. 2017; Camacho et al. 2018; Steele et al. 2019). Treating window counts as a census proved beneficial as being an easy, straight-forward method ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (Joint Columbia River Management Staff 2019).

However, using window counts as a census of fish passing LGR can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes. Typically, the observational error rates of sp/sum Chinook Salmon and steelhead counted passing a window on the fish ladder, using both in-person and video methods, are unknown, and sampling error rates are ignored. Hatch et al. (1994) conducted a study of the counting system at LGR, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters when compared to video counts. There was also a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; the undercounting was attributed to a significant nighttime passage rate (crossing while the counting window was closed) with an estimated 3.5\% of sp/sum Chinook Salmon and 6.6\% of steelhead adults passing during ``nighttime'' hours. They also found some species misidentification issues, which could result in either under- or over-counting of a particular species. While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, and at a time when sp/sum Chinook Salmon and steelhead total returns were typically lower than in the past decade.

One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement of these species at LGR. Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again (Boggs et al. 2004) and be counted multiple times at the window (re-ascension). Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement (Dauble and Mueller 2000). Previously, it has often been assumed that these two biological processes essentially cancel each other out, resulting in window counts that provide an unbiased estimate of escapement (Camacho et al. 2018), which may be true in many or most years, but perhaps not always.

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detections of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach (Royle and Dorazio 2008) which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are divided into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Our model is implemented in the \textbf{ST}ate-space \textbf{A}dult \textbf{D}am \textbf{E}scapement \textbf{M}odel (STADEM) package for the statistical software R (R Core Team 2020), and is available for download or installation from \url{https://github.com/KevinSee/STADEM}. To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated ``truth''. Then, as a case study, we applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.

@@

Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other species passing LGR (e.g., fall Chinook Salmon) or other fish passage facilities (e.g., Bonneville Dam).

@@

Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the LGR fish ladder were compiled and provided by the US Army Corps of Engineers. When summed, they provide an estimate of the number of fish ascending and passing LGR each season. Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours (Hatch et al. 1994). Direct visual monitoring occurred during peak run times (April 1 -- October 31) for 16 hours per day (0400 -- 2000 hours). Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 -- March 31 and November 1 -- December 31) for 10 hours per day (0600 -- 1600 hours) (USACE 2015). During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes each hour when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) (USACE 2015). Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, \url{www.cbr.washington.edu/dart/query/adult_daily}, using their window count query. Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2.

@@

The second source of data are from a sample of fish collected in the adult trap as they migrated past LGR (Ogden 2016a). The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., phenotypic origin {[}wild, hatchery{]}, tissue sample for genetic data, length, scale sample for age, sex) from captured adults which allows for decomposition of the escapement into specific groups (e.g., Camacho et al. 2018; Steinhorst et al. 2017). The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns. Trap sample rates were typically 10-25\%, but fluctuated throughout the season due to, for example, high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, or adjustments due to shifting species composition throughout the year.

All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length, and visually identified as wild or hatchery. The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are also used in less than 10\% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips. For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age (Wright et al. 2015) and genetic tissue samples were used to determine sex (Campbell et al. 2012) and estimate the location of origin of wild fish using genetic stock identification (e.g., Hargrove et al. 2019). Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and to better estimate the true proportion of unclipped hatchery fish that can otherwise appear phenotypically wild. Camacho et al. (2018) provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging based on the genetic samples taken (Steele et al. 2013, 2019). Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game (Camacho et al. 2018).

\hypertarget{pit-tag-data}{%
\subsubsection{PIT Tag Data}\label{pit-tag-data}}

The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at PIT tag detection sites located in the LGR fish ladder. These observations provide data related to 1) an estimate of the ``true'' trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate. Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at LGR (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data were provided through DART and the adult ladder PIT tag query; \url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}.

@@

\hypertarget{total-and-weekly-escapement}{%
\subsubsection{Total and Weekly Escapement}\label{total-and-weekly-escapement}}

Escapement at LGR is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model (Royle and Dorazio 2008), with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGR each week. We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified. The log of the true number of fish crossing during time-step \(t\), (\(X_t\)), is modeled as a random walk process (Shumway and Stoffer 2010), with \(e_t\) representing the process errors drawn from a normal distribution with variance \(\sigma^2_X\).

\begin{equation}
  \begin{aligned}
    \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
    e_t &\sim \text{Normal}(0, \sigma^2_X)
  \end{aligned}
\end{equation}
